---
title: "5. Descriptive Statistics"
author: "Akihito Kamata"
date: "September 22, 2020"
output: html_notebook
---

## Load Required Packages
```{r message=F}
library(tidyverse)
library(sjlabelled)
library(sjmisc)
library(psych)
```

## Read a Data file
Read `easyCBM.sav` file.  
As we have seen before, this data file contains 15 variables for randomly selected 250 students, longitudinal easyCBM test scores and demographic variables.

`ID`: Deidentified ID.  
`ORF1`: Oral reading fluency score in Fall.  
`ORF2`: Oral reading fluency score in Winter.  
`ORF3`: Oral reading fluency score in Spring.  
`MCC1`: Reading comprehension score in Fall.  
`MCC2`: Reading comprehension score in Winter.  
`MCC3`: Reading comprehension score in Spring.  
`VOC1`: Vocabulary score in Fall.  
`VOC3`: Vocabulary score in Spring.  
`Test_A`: Statewide test score.  
`Ethnic`: Ethnicity (1=Amer Ind/Alsk Nat, 2=Asian/Pac Isl, 3=Black, 4=Hispanic, 5=White, 6=Multi-Ethnic, 7=Decline).  
`EconDs`: Economically Disadvantaged (0=No, 1=Yes).  
`SpEd`: Special Education (0=No, 1=Yes).  
`Gender`: Gender (0=Males, 1=Females).  
`Prof`: Reading proficiency status (0=Not Meet, 1=Meets/Exceeds).  

```{r}
easyCBM10 <- read_spss("easyCBM.sav")
```

## Descriptive Statistics
There are MANY different ways to generate descriptive statistics. For this class, I recommend you get used to three different ways.

### (1) descr() function in the `sjmisc` package
This is my favorite at this point. It allows us to customize output very well. Also, the output is very easy to extract for some other purposes.  

Default specification, which shows all available statistics.
```{r}
easyCBM10 %>% 
  select(MCC1, MCC2, MCC3) %>% 
  descr()
```
Specify a short version of the output, which shows a minimal set of statistics.
```{r}
easyCBM10 %>% 
  select(MCC1, MCC2, MCC3) %>% 
  descr(show = "short")
```
We can specify a custom set of statistics.
```{r}
easyCBM10 %>% 
  select(MCC1, MCC2, MCC3) %>% 
  descr(show = c("label", "n", "mean", "sd", "skew"))
```
Customize the output format.
```{r}
easyCBM10 %>% 
  select(MCC1) %>% 
  descr(show = c("label", "n", "mean", "sd", "skew"),
        out = "viewer")
```

### (2) describe() function in the `psych` package
This used to be my favorite. It has two potential advantage. 1) It computes kurtosis, which `descr()` function does not, and 2) By default, it produces an interactive table for R Notebook file, both for Rmd and HTML file. However, it does not come with flexibility of what `descr()` can do.  

Default specification, which shows all available statistics.  
```{r}
easyCBM10 %>% 
  select(MCC1, MCC2, MCC3) %>% 
  describe()
```
Shorter version of output.
```{r}
easyCBM10 %>% 
  select(MCC1, MCC2, MCC3) %>% 
  describe(fast = T)
```

### (3) `summary()` function as part of the base packages
This is quite limited. However, it produces 1st and 3rd quartiles, which the other two do not produce.  

```{r}
easyCBM10 %>% 
  select(MCC1, MCC2, MCC3) %>% 
  summary()
```

## Grouped Statistics
We can use `group_by()` function to generate grouped statistics.  

Let's say we are interested in descriptive statistics of `MCC1`, `MCC2`, and `MCC3`, grouped by `SpEd` groups.  

First, let's generate a frequency table for `SpEd`.  
```{r}
easyCBM10 %>% select(SpEd) %>% frq()
```

Grouped statistics now by using `group_by()` function.
```{r}
easyCBM10 %>% 
  group_by(SpEd) %>% 
  select(MCC1, MCC2, MCC3) %>% 
  descr(show = c("label", "n", "mean", "sd"))
```

## Contingency Table
If we describe grouped frequencies, we are dealing with what is called a "contingency table". Technically, when we describe the frequencies by crossing two or more categorical variables, the table we produce is called a contingency table.  

For example, let's say we are interested in the frequencies of `Gender` grouped by `SpEd`. In other words, we will cross two categorical variables `SpEd` and `Gender`. As a result, we will obtain frequencies of the following four groups:  

SpEd = No & Gender = Male
SpEd = No & Gender = Female
SpEd = Yes & Gender = Male
SpEd = Yes & Gender = Female

```{r}
easyCBM10 %>% 
  group_by(SpEd) %>% 
  select(SpEd, Gender) %>% 
  frq()
```

If we prefer a simpler table only with frequencies, we can use `flat_table()` function in the `sjmisc` package, instead of `frq()` function.

```{r}
easyCBM10 %>% 
  group_by(SpEd) %>% 
  select(SpEd, Gender) %>% 
  flat_table()
```

